---
title: "Week 2 Exercises"
author: "Melissa Nooney"
date: "July 6, 2024"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

Please complete all exercises below. You may use stringr, lubridate, or the forcats library. 

Place this at the top of your script:
```{r}
library(stringr)
library(lubridate)
library(forcats)
library(tinytex)
```



# Exercise 1
Read the sales_pipe.txt file into an R data frame as sales. 

```{r}
# Your code here
sales <- read.delim("sales_pipe.txt"
                     ,stringsAsFactors=FALSE
                     ,sep = "|"
 )
```

# Exercise 2
You can extract a vector of columns names from a data frame using the colnames() function. Notice the first column has some odd characters. Change the column name for the FIRST column in the sales date frame to Row.ID. 

**Note: You will need to assign the first element of colnames to a single character.**

```{r}
colnames(sales)[1] <- 'row.id'

```

# Exercise 3
Convert both Ship.Date and Order.Date to date vectors within the sales data frame. What is the number of days between the most recent order and the oldest order?
How many years is that?
How many weeks?

**Note: Use lubridate**

```{r}
sales$Order.Date <- as.Date(sales$Order.Date
                                ,format='%m/%d/%Y'
                               )

sales$Ship.Date <- mdy(sales$Ship.Date)

class(sales$Ship.Date)

day_min <- min(sales$Order.Date)
day_max <- max(sales$Order.Date)
days_between <- day_max-day_min
days_between
weeks_btwn <- difftime(day_max, day_min, units = "weeks")
weeks_btwn
years<- interval(day_min, day_max) / years(1)
years
```


# Exercise 4
What is the average number of days it takes to ship an order?

#reviewing my answers, not sure why I created a new column, I could have also 
just applied mean(shipping), but I'm going to leave it. I do realize I could 
have made this much easier on myself. 
```{r}
shipping <- sales$Ship.Date - sales$Order.Date
shipping
sales$ship.days <- shipping
mean(sales$ship.days)

```


# Exercise 5
How many customers have the first name Bill?
You will need to split the customer name into first and last name segments and then use a regular expression to match the first name bill. Use the length() function to determine the number of customers with the first name Bill in the sales data. 

```{r}
temp_name <- str_split_fixed(sales$Customer.Name, pattern=' ',n=2)
Bill <- which(temp_name == "Bill")
length(Bill)
```

# Exercise 6
How many mentions of the word 'table' are there in the Product.Name column? 
**Note you can do this in one line of code**

```{r}
sum(str_detect(sales$Product.Name, "table"))
```

# Exercise 7
Create a table of counts for each state in the sales data. The counts table should be ordered alphabetically from A to Z. 

#I struggled so badly with this problem. In an attempt to get my mind off it,
I actually found what I believe to be the correct code, by doing the discussion
assignment! And it was painfully simple.
```{r}
table(sales$State)


```

# Exercise 8
Create an alphabetically ordered barplot for each sales Category in the State of Texas. 

```{r}
texas <- subset(sales, subset = State == "Texas")
barplot(table(texas$Category))

```

# Exercise 9
Find the average profit by region.
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r}
aggregate(sales$Profit, list(sales$Region), mean)

```

# Exercise 10
Find the average profit by order year. 
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r}
temp_year <- str_split_fixed(sales$Order.Date, pattern='-',n=3)
sales$year <- temp_year[,-2:-3]
aggregate(sales$Profit, list(sales$year), mean)

```

