---
title: "DataFrames_and_Lists"
author: "HDS"
date: "2024-06-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Frames and Lists

A dataframe in R is pretty much an R version of an SQL data table

Modern computer languages borrow concepts from one another constantly.

Python has a library called Pandas that implements a dataframe in Python that 
works me in a very similar way.

In a dataframe (or SQL data table) each row is an "observation" or an "event" 
an "individual item".   

All the variable values on a given row belong to the same observation

The columns of a dataframe are different variables, one per column.   The 
columns do not have to be the same type of data as the other columns.

Dataframes look much like Excel sheets, but the structure is more rigid.

Typically you will be loading a dataframe from some data source, maybe a 
database, a website or a table created in Excel.

Right now we will look at an example data set that comes with R the "Motor Trend
cars" data set mtcars.   This is a built in "lab rat" data set that comes with
R

We will load the data first

```{R}
data(mtcars)
```

We can find out what is in it using STR

```{R}
str(mtcars)
```
Like a matrix, the data frame has rows and columns we can index by number

There are 32 rows here and 11 columns

We can look at the first 6 rows of the data frame, this is a nice way to 
see what is in a data frame

```{R}
head(mtcars)
```

We can see the last 6 lines using the tail() function, or the whole
dataframe using the View() function

It is a bad idea to depend too heavily on the View() option.

This is a small data set, 32 rows by 11 columns, so it is easy to understand
what is happening, but when you get into the hundreds or thousands of rows,
View() is really useful anymore

```{R}
View(mtcars)
```

On the other hand, I use str(), head() and colnames() constantly

We can get lists of the row and column names

```{R}
rownames(mtcars)
```

```{R}
colnames(mtcars)
```

#Summary

Most R variables will show you some interesting information when you
use the summary() function

```{R}
summary(mtcars)
```


#indexing a data frame

Indexing works the same way it does for matrices or vectors

Index with integers

```{R}
mtcars[1,1:5]
```


```{R}
mtcars[2,]
```

```{R}
mtcars[,10]
```

```{r}
mtcars["Mazda RX4","hp"]

```

#Working with a column

We can work with a column by giving the name of the dataframe, $ and the column
name

```{R}
mtcars$wt
```

#Adding a column

R does not "protect" data frames, we can just add a new column by 
giving it a column name and a set of values- this is a bit alarming if you
are used to managing SQL databases :)

Example:

In cars, the horsepower to Weight ratio is important (it is for humans too)

We can add this variable to our data frame

```{R}
mtcars["Power_2_Weight"]=mtcars$hp/mtcars$wt
```

We can now ask, which car has the highest power to weight ratio?

```{R}
maxPW_index=which.max(mtcars$Power_2_Weight)

mtcars[maxPW_index,]
```

It might be sort of interesting to see the top three power to weight vehicles,
not just the largest.

We'd like to sort the data frame by Power_2_weight ratio

We use order to get the indexing to do this

```{R}
p2w_index=order(mtcars$Power_2_Weight,decreasing = TRUE)

head(mtcars[p2w_index,])

```

Look at those ideas- think of how you might use them on your own problem
to work with data

#Plotting

We will see more sophisticated plotting, but we can easily do simple
plots using the plot() or hist() functions

```{R}
plot(mtcars$Power_2_Weight,mtcars$qsec)
```
 What hypothesis does the plot above address?
 
#Histograms are one way to look at distributions of values

```{r}
hist(mtcars$Power_2_Weight)
```
#Question/Action

State a hypothesis about Power_2_Weight and mpg (miles per gallon).   Create
a graph that tests the hypothesis

#Creating data frames in R

We can create data frames from within R in several ways

Lets declare some matched arrays, maybe describing some people

```{R}
last_name=c("Smith","Jones","Alvarez")
first_name=c("Bob","Jane","Angel")
age=c(30,27,32)
```

We can put these in a dataframe

```{R}
customer_df=data.frame(last=last_name, first=first_name,Age=age)
head(customer_df)
```

#Utilities

There are tools for reading data out of a csv (comma seperated value) or excel
file and into a data frame,  or to connect to a database and convert an SQL
data table into a data frame.

There are also tools to write a data file to disk

It is often convenient to save results as data.frames, since they are easy to
work with in other software packages if you save them as csv files

#Loading a DataFrame from a File

We need to specify the name of the file, with the file path included.

The easiest way to get the full file path correct is to use the
file.choose() function to browse for the file.

Here,  we browse for the file and save it to a variable called "infile"

One thing to be careful about is that file.choose() will not knit,
use it in your work, but copy the file name to a variable and then 
"rem out" the file.choose() operation by putting # in front of it


We will use the Boston Assessment Roll again

```{R}

#rem this out by putting # in front of the command below

file.choose()
```

```{R}
# cut and paste using Ctrl-C and Ctrl-V to assign the full path to infile

infile="C:\\Users\\hdavi\\Dropbox\\Merrimack_Data_Science\\DSE5002_R+Python\\DSE5002_Module_1\\Boston_Assessment_Roll_2024.csv"
```

now use read.csv() to load the assessment roll into a data frame

```{R}
boston_roll=read.csv(infile)
```

```{R}
head(boston_roll)
```

There are R functions to import excel files and many other types of data

Google "R import excel sheet" to learn how to do this

#Saving a dataframe to disk

This will let us use it somewhere else later

We do have to specify the whole path name,  meaning the directory name
plus the file name.

If you don't specify a full path,  R saves the file to the current directory

YOu can see the current directory in the Files window to the right in 
RStudio

you can see the current directory using getwd(), which means get working 
directory

```{R}
getwd()
```

You can change the directory using the "Session" menu in RStudio

I'll change to the desktop using Session-Set Working Directory

```{R}
getwd()
```

Let's save our customer_df data frame

specify the file name

```{r}
#I like to put the date in the name of an output file, it helps in long 
#projects

outfile="customer_df_6_21_2024.csv"
```

```{R}
write.csv(customer_df,outfile)
```

Open this file in excel and see if it matches what you expect

#Lists

R can store things in lists, which can contain different types of objects

```{R}
a=list("Bob", 1, as.integer(2),TRUE,c(1,3,5,7,9))
```

Lists can store different types of things

We can index them

```{R}
a[1]
```
we can assign names in a list as well

```{R}
a=list(name="Bob", purchases=1, category=as.integer(2),active=TRUE,items_purchased=c(1,3,5,7,9))
a
```
We can access these by name

```{R}
a$name
```

Notice that one of the things in a,  items_purchased, is itself a vector.

You cannot store vector values within a single variable such as 
"items_purchased" in a data frame,  

You can't store this easily in a single SQL dataframe either

You can in a newer NOSQL database like MongoDB, or in a user defined class in R

More on that later....

We can't really do math directly on a list, since they often have mixtures of
data types in them.

Right now,  dataframes will be more usefull to you

If you get working on complex problems, or start using NOSQL databases, you will
be using lists more often.



